---
title: "Project_251"
format: pdf
author: "Brandon Upper, Ben Murdock, Hannah Lyons"
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

```{r}
#library
library(vroom)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(patchwork)
```

```{r}
#data
shootouts <- vroom('./WorldCupShootouts.csv')
```

## Introduction:

*(Explain your research question (and necessary background information)
as well as the value of answering it.)* 

Penalty shootouts are one of the
most intense moments in soccer, and every choice a player makes can
influence the outcome. One detail that often gets overlooked is whether
a player is left-footed or right-footed and if that affects the chance
of scoring. Our research question asks if there is a difference between
the proportion of successful penalty kicks taken by right-footed players
compared to left-footed players. Understanding this matters because many
coaching decisions during a shootout are based on patterns and
probabilities. If there is an advantage connected to success, and in
this case dominant feet, even a small one could influence which players
a coach chooses for these high-pressure moments. It can also help us
understand how kickers approach the goal and whether their dominant foot
provides any advantages

## Methodology:

*(Explain how the data were obtained, as well as providing a citation of
their source (if available). Provide a summary of the data by reporting
summary statistics and appropriate figures. Identify the observational
unit, select a data model (likelihood) and justify this choice (i.e.,
determine if assumptions are reasonable and if not, provide some
possible consequences). Explain what the relevant parameters of the
likelihood are and how they will be useful in answering your question.
Choose and justify your prior distributions. )*

Our dataset contains information on World Cup penalty shootouts from
Spain 1982 to Russia 2018 sourced from Kaggle. URL: [World Cup Penalty
Shootouts
Dataset](https://www.kaggle.com/datasets/pablollanderos33/world-cup-penalty-shootouts)

Each observation represents one penalty kick. The dataset contains 305
rows in total. The first row is a descriptive header, leaving 304 rows
of actual data and 9 columns. The columns include:

-   **Game_id:** Each team in a single game gets a number. Numbers are
    not repeated across games.
-   **Team:** The country team of the kicker.
-   **Zone:** The goal was divided into 9 sections, each labeled by
    number. This number is the zone.
-   **Foot:** The foot used to take the shot (Left/Right).
-   **Keeper:** The direction the goalkeeper dove (Left/Right).
-   **On Target:** Whether the shot would have been a goal if the keeper
    had not intervened (1 = Yes, 0 = No).
-   **Goal:** Whether the penalty was successfully scored (1 = Goal, 0 =
    Miss)
-   **Penalty number:** The order of the penalty within the shootout
-   **Elimination:** Indicates whether the penalty could end the game (1
    = Yes, 0 = No)

Since we are only interested in whether a shot was scored or missed, our
key variables are Foot and Goal.There is a lot more right footed people
who took penalties, likely because they are more common in the general
population, but interestingly they have a lower ratio of success. This
is something to consider when "determining if there is a difference
between proportion of successful penalty kicks of right footed vs. left
footed soccer players." There seems to be no strong correlations between
any of the numeric variables.

This is useful as we can see that this data is unlikely to have issues
with multicollinearity. It's clear that different zones have higher
portions of goals scored, and when we recognize that the zone targeted
and thereby goal likelihood can be highly influenced by right vs left
footed kickers. From prior soccer experience, we could look for
tendencies in the final project for kickers to shoot towards met
opposite of their dominant foot. The second graph also displays how long
the shootout lasts for, and its clear that most of them end before or
around the twelfth number.

For our likelihood, we treat each penalty kick as a Bernoulli trial
because the outcome is binary. A kick is either a success(1) or a
failure(0). We assume that the kicks are independent within
right/left-foot groups. Therefore, each group has its own probability of
scoring, and these are the parameters we want to estimate. We chose a
Beta(1, 1) prior for both groups, which is uniform because we do not
have strong prior beliefs about which group should be more successful.
Using an uninformed Beta(1, 1) prior lets the data speak for itself, but
it can also make our results more sensitive when one group has a small
sample size. This means the posterior for left footed players may be
wider and less stable because we are not adding any prior information to
help anchor the estimate.

## Results:

*(Perform posterior inference on the function of the parameters that
directly answers your primary research question (for example, the
difference in means, or the ratio of variances). Report and properly
interpret an associated credible interval for this function of
parameters. Include properly formatted and informative graphic(s) for
this function of parameters with its corresponding prior.)*


```{r}

# Change directory
setwd("~/GitHub/Penalty-Kick-Project")

# Pull data and remove columns that aren't foot and goal
shootouts <- vroom('WorldCupShootouts.csv')
shootouts <- shootouts %>% select(c("Foot","Goal"))
  
Right_Foot_Data <- shootouts %>% filter(Foot == "R")
Left_Foot_Data <- shootouts %>% filter(Foot == "L")

# Computing the Y and N values for Right and Left based on data

Right_Foot_Y <- sum(pull(Right_Foot_Data, Goal))
Left_Foot_Y <- sum(pull(Left_Foot_Data, Goal))

Right_Foot_N <- length(pull(Right_Foot_Data, Goal))
Left_Foot_N <- length(pull(Left_Foot_Data, Goal))

# Prior which is uninformative in our case

alpha <- 1
beta  <- 1 

prior_R <- rbeta(10000, alpha, beta)
prior_L <- rbeta(10000, alpha, beta)
prior_diff <- prior_R - prior_L

# right foot a star and b star

a_star_R <- alpha + Right_Foot_Y
b_star_R <- beta  + Right_Foot_N - Right_Foot_Y

# left foot  a star and b star
a_star_L <- alpha + Left_Foot_Y
b_star_L <- beta  + Left_Foot_N - Left_Foot_Y


set.seed(43110)
post_R <- rbeta(1000, a_star_R, b_star_R)
post_L <- rbeta(1000, a_star_L, b_star_L)

Combined_df <- data.frame(Difference = post_R - post_L)

# Prior Plot
Prior_df <- data.frame(Difference = prior_diff)

G1 <- ggplot(Prior_df, aes(x = Difference)) +
  geom_histogram(fill = "darkgreen", color = "black") +
  labs(title = "Prior Distribution of p_right - p_left",
  x = "Difference in Success Proportion",
  y = "Count")


# Posterior Plot
G2 <- ggplot(Combined_df, aes(x = Difference)) +
  geom_histogram(fill = "darkred", color = "black") +
  labs(title = "Posterior Distribution of p_right - p_left",
  x = "Difference in Success Proportion",
  y = "Count")

# Patchwork Plot Both at once
G1 + G2

# Calculate 95% Credible Interval
R_L <- post_R - post_L
quantile(R_L, c(0.025, 0.975))

```


```{r}
library(tidyverse)

#Load the data
shootouts <- vroom('WorldCupShootouts.csv')

clean_data <- shootouts %>%
  filter(Foot %in% c("L", "R"), !is.na(Goal)) %>%
  mutate(Foot = factor(Foot, levels = c("L", "R"), labels = c("Left", "Right")))

#Prior Params
# Prior: Beta(1, 1)
prior_alpha <- 1
prior_beta <- 1

# Get counts
counts <- clean_data %>%
  group_by(Foot) %>%
  summarise(
    successes = sum(Goal),
    n = n()
  )

y_L <- counts$successes[counts$Foot == "Left"]
n_L <- counts$n[counts$Foot == "Left"]
y_R <- counts$successes[counts$Foot == "Right"]
n_R <- counts$n[counts$Foot == "Right"]

#Posterior Params
alpha_L <- prior_alpha + y_L
beta_L  <- prior_beta + (n_L - y_L)

alpha_R <- prior_alpha + y_R
beta_R  <- prior_beta + (n_R - y_R)

# Mean = alpha / (alpha + beta)
# Var  = (alpha * beta) / ( (alpha + beta)^2 * (alpha + beta + 1) )

mu_L <- alpha_L / (alpha_L + beta_L)
var_L <- (alpha_L * beta_L) / ((alpha_L + beta_L)^2 * (alpha_L + beta_L + 1))

mu_R <- alpha_R / (alpha_R + beta_R)
var_R <- (alpha_R * beta_R) / ((alpha_R + beta_R)^2 * (alpha_R + beta_R + 1))

# Diff ~ Normal(mean_diff, sd_diff)
mean_diff <- mu_L - mu_R
var_diff  <- var_L + var_R
sd_diff   <- sqrt(var_diff)

# Calculate 95% Credible Interval
ci_lower <- qnorm(0.025, mean = mean_diff, sd = sd_diff)
ci_upper <- qnorm(0.975, mean = mean_diff, sd = sd_diff)

# Calculate Probability Left > Right
prob_L_better <- 1 - pnorm(0, mean = mean_diff, sd = sd_diff)

#VISUALIZATION
x_vals <- seq(mean_diff - 4*sd_diff, mean_diff + 4*sd_diff, length.out = 1000)
y_vals <- dnorm(x_vals, mean = mean_diff, sd = sd_diff)
plot_df <- data.frame(Difference = x_vals, Density = y_vals)

ggplot(plot_df, aes(x = Difference, y = Density)) +
  geom_line(color = "lightblue", size = 1.2) +
  geom_area(fill = "lightblue", alpha = 0.3) +
  # Add CI
  geom_vline(xintercept = c(ci_lower, ci_upper), linetype = "dashed") +
  annotate("text", x = ci_upper + 0.02, y = max(y_vals)/2, label = "95% CI", angle = 90) +
  # Add referance line at zero
  geom_vline(xintercept = 0, color = "red", size = 0.5) +
  labs(
    title = "Posterior Difference of Left vs Right Success",
    subtitle = paste("95% CI:", round(ci_lower, 3), "to", round(ci_upper, 3)),
    x = "Difference in Proportion (Left - Right)",
    y = "Density"
  ) +
  theme_minimal()
```


## Discussion:

*(Instructions: Summarize your results, identify shortcomings of the
work, and explain what could be done in the future to either better
answer your question or to address a natural follow-up question.)*

## Appendix:

*(include all R code necessary to conduct your analysis.)*

The proposed data source and a well-organized and well-articulated
exploratory analysis of its contents (this should include well-labeled
plots created in R).

Libraries

```{r}
library(ggplot2)
library(vroom)
library(tidyverse)
```

Read File

```{r}
Soccer_data <- vroom("WorldCupShootouts.csv")
summary(Soccer_data)
head(Soccer_data)
```

Reading in our file we can see various factors, features and variables
that are described more in depth in the above short description of the
data. From this code we can see the first five rows of the data and
notice that many points are read in as doubles and characters. For the
final analysis, we likely will convert some of these variables for
simplicity

Basic plots about Goals and Penalties

```{r}
Soccer_data <- Soccer_data %>%
  filter(is.finite(Zone)) 
ggplot(data = Soccer_data) + 
  geom_bar(mapping = aes(x = Zone), 
           fill = "lightblue") + ggtitle("Goals by Zone") + ylab("Count")
ggplot(data = Soccer_data) + 
  geom_bar(mapping = aes(y = Penalty_Number), fill = "navyblue") + 
  ggtitle("Penalty Number by Count") + ylab("Penalty Number") + xlab("Count")
```

These graphs give some visuals to some of the data and the types
penalties we're seeing. For example, its clear that different zones have
higher portions of goals scored, and when we recognize that the zone
targeted and thereby goal likelihood can be highly influenced by right
vs left footed kickers. From prior soccer experience, we could look for
tendencies in the final project for kickers to shoot towards met
opposite of their dominant foot. The second graph also displays how long
the shootout lasts for, and its clear that most of them end before or
around the twelfth number.

Looking for Correlation

```{r}
numeric_data <- Soccer_data %>%
  select(where(is.numeric))

library(corrplot)
cor_matrix <- cor(numeric_data)
corrplot(cor_matrix)
```

There seems to be no strong correlations between any of the numeric
variables. This is useful as we can see that this data is unlikely to
have issues with multicollinearity.

Left vs right foot - success proportion

```{r}
#

plot_data <- Soccer_data %>%
  select(Foot, Goal) %>% 
  filter(Foot %in% c("L", "R"),
         Goal %in% c(0, 1)) %>%
  mutate(Goal_Status = factor(Goal, 
                              levels = c(0, 1), 
                              labels = c("Miss", "Success")))

ggplot(plot_data, aes(x = Foot, fill = Goal_Status)) +
  geom_bar(position = "dodge")

table(plot_data)
print("Success rate L = 0.3214286 vs R = 0.2959641")
```

Some basic data visualization looking at the success rate of left foot
vs right foot to see if one generally has a higher rate. There is a lot
more right footed people who took penalties, likely because they are
more common in the general population, but interestingly they have a
lower ratio of success. This is something to consider when "determining
if there is a difference between proportion of successful penalty kicks
of right footed vs. left footed soccer players."
