---
title: "Project_251"
format: pdf
author: "Brandon Upper, Ben Murdock, Hannah Lyons"
---

# Hannah: Short Written Description + Extra Graph

Also include a short written description of the data set (including a URL if possible).

# Ben: Proposed Research Question

Your proposed research question, why you chose that research question, and how the data will permit you to answer. Left vs Right foot.

We would like to determine if there is a difference between number of successful penalty kicks of right footed vs. left footed soccer players. We will measure the rate in proportion of successes kicks.

# Brandon: EDA

Libraries

```{r}
library(ggplot2)
library(vroom)
library(tidyverse)
```

Read File

```{r}
Soccer_data <- vroom("WorldCupShootouts.csv")
summary(Soccer_data)
head(Soccer_data)
```

Basic plots about Goals and Penalties

```{r}
Soccer_data <- Soccer_data %>%
  filter(is.finite(Zone)) 
ggplot(data = Soccer_data) + geom_bar(mapping = aes(x = Zone), fill = "lightblue") 
ggplot(data = Soccer_data) + geom_bar(mapping = aes(y = Penalty_Number), fill = "navyblue")
```

Looking for Correlation

```{r}
numeric_data <- Soccer_data %>%
  select(where(is.numeric))

library(corrplot)
cor_matrix <- cor(numeric_data)
corrplot(cor_matrix)
```
