---
title: "Project_251"
format: pdf
author: "Brandon Upper, Ben Murdock, Hannah Lyons"
---
```{r}
#library
library(vroom)
library(ggplot2)
library(tidyverse)
library(dplyr)
```

```{r}
#data
shootouts <- vroom('./WorldCupShootouts.csv')
```


# Hannah: Short Written Description + Extra Graph
Also include a short written description of the data set
(including a URL if possible).  



# Ben: Proposed Research Question, left/right success bar graph.

We would like to determine if there is a difference between proportion of 
successful penalty kicks of right footed vs. left footed soccer players.



# Brandon: EDA

The proposed data source and a well-organized and well-articulated exploratory 
analysis of its contents (this should include well-labeled plots created in R).

```{r}
#left vs right foot

plot_data <- shootouts %>%
  select(Foot, Goal) %>% 
  filter(Foot %in% c("L", "R"),
         Goal %in% c(0, 1)) %>%
  mutate(Goal_Status = factor(Goal, 
                              levels = c(0, 1), 
                              labels = c("Miss", "Success")))

ggplot(plot_data, aes(x = Foot, fill = Goal_Status)) +
  geom_bar(position = "dodge")
```



