---
title: "Project_251"
format: pdf
author: "Brandon Upper, Ben Murdock, Hannah Lyons"
---

```{r}
#library
library(vroom)
library(ggplot2)
library(tidyverse)
library(dplyr)
```

```{r}
#data
shootouts <- vroom('./WorldCupShootouts.csv')
```

#Short Written Description Our dataset contains information on World Cup penalty shootouts from Spain 1982 to Russia 2018 sourced from Kaggle. URL: [World Cup Penalty Shootouts Dataset](https://www.kaggle.com/datasets/pablollanderos33/world-cup-penalty-shootouts)

The dataset contains 305 rows in total. The first row is a descriptive header, leaving 304 rows of actual data and 9 columns. The columns include: Game_id: Each team in a single game get's a number. Numbers are not repeated across separate games. Team: The country team of the kicker. Zone: The goal was divided into 9 sections, each labeled by number. This number is the zone. Foot: The foot used to take the shot (Left/Right). Keeper: The direction the goalkeeper dove(Left/Right) On Target: Indicates whether the shot would have been a goal if the keeper had not intervened(1 = Yes, 0 = No) Goal: Whether the penalty was successfully scored (1 = Goal, 0 = Miss) Penalty number: The order of the penalty within the shootout Elimination: Indicates whether the penalty could end the game (1 = Yes, 0 = No)

# Ben: Proposed Research Question, left/right success bar graph.

We would like to determine if there is a difference between proportion of successful penalty kicks of right footed vs. left footed soccer players.

# Brandon: EDA

The proposed data source and a well-organized and well-articulated exploratory analysis of its contents (this should include well-labeled plots created in R).

```{r}
#left vs right foot - success proportion

plot_data <- shootouts %>%
  select(Foot, Goal) %>% 
  filter(Foot %in% c("L", "R"),
         Goal %in% c(0, 1)) %>%
  mutate(Goal_Status = factor(Goal, 
                              levels = c(0, 1), 
                              labels = c("Miss", "Success")))

ggplot(plot_data, aes(x = Foot, fill = Goal_Status)) +
  geom_bar(position = "dodge")
```
